---
title: "Out of Area Placements Birmingham"
author: "Vanessa Fillis"
date: "08/01/2021"
output: html_document
---

# Out of Area Placements in Mental Health Services in the Birmingham and Solihull Mental Health NHS Foundation Trust

## Activate packages

```{r}
library(lubridate)
library(tidyr)
library(dplyr)
```

## Data import

In a previous step I have downloaded data for Out of Area Placements (OAPs) in mental health services between October 2016 and October 2020 (latest available data set) from [NHS Digital](https://digital.nhs.uk/data-and-information/publications/statistical/out-of-area-placements-in-mental-health-services).

```{r}
OAPs <- read.csv("oaps2016to2020.csv", stringsAsFactors = FALSE)
```

## Inspect the data frame

What I can see:

- The column containing the date is not in date format, but as character. I have to add a date column that contains the publication periodm in date format. Because the data frame is so large I will to this *after* I have filtered for the questions I want to look at. 
- The column **value** is in character format as well. I will transform it into numeric values. 

```{r}
head(OAPs)
OAPs$value <- as.numeric(OAPs$value)
head(OAPs)
```

## Filter for total inappropriate OAP days in Birmingham and Solihull Mental Health NHS Foundation Trust for each period

```{r}
OAPs.trusts.days <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.trusts.days <- subset(OAPs.trusts.days, OAPs.trusts.days$Question == "Total number of inappropriate OAP days over the period" | OAPs.trusts.days$Question == "Total number of inappropriate out of area placement days over the period")

OAPs.trusts.days <- subset(OAPs.trusts.days, OAPs.trusts.days$Breakdown1 == "SendingProvider")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.trusts.days <- tidyr::separate(data = OAPs.trusts.days, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.trusts.days$beginning.of.publication.period <- lubridate::dmy(OAPs.trusts.days$beginning.of.publication.period)

OAPs.trusts.days$end.of.publication.period <- lubridate::dmy(OAPs.trusts.days$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.trusts.days <- OAPs.trusts.days[-c(9:11)]
```

### Filter for Birmingham and Solihull Mental Health NHS Foundation Trust

```{r}
OAPs.bham.days <- subset(OAPs.trusts.days, OAPs.trusts.days$Breakdown1Description == "BIRMINGHAM AND SOLIHULL MENTAL HEALTH NHS FOUNDATION TRUST" | OAPs.trusts.days$Breakdown1Description == "Birmingham And Solihull Mental Health NHS Foundation Trust")
```

### Export

```{r}
write.csv(OAPs.bham.days, file = "OAPs_bham_days.csv")
```

## Filter for costs caused by inappropriate OAP in Birmingham and Solihull Mental Health NHS Foundation Trust for each period

```{r}
OAPs.trusts.costs <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.trusts.costs <- subset(OAPs.trusts.costs, OAPs.trusts.costs$Question == "Total Cost For Inappropriate OAPs in Period" | OAPs.trusts.costs$Question == "Total Cost For Inappropriate out of area placements in Period")

OAPs.trusts.costs <- subset(OAPs.trusts.costs, OAPs.trusts.costs$Breakdown1 == "SendingProvider")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.trusts.costs <- tidyr::separate(data = OAPs.trusts.costs, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.trusts.costs$beginning.of.publication.period <- lubridate::dmy(OAPs.trusts.costs$beginning.of.publication.period)

OAPs.trusts.costs$end.of.publication.period <- lubridate::dmy(OAPs.trusts.costs$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.trusts.costs <- OAPs.trusts.costs[-c(9:11)]
```

### Filter for Birmingham and Solihull Mental Health NHS Foundation Trust

```{r}
OAPs.bham.costs <- subset(OAPs.trusts.costs, OAPs.trusts.costs$Breakdown1Description == "BIRMINGHAM AND SOLIHULL MENTAL HEALTH NHS FOUNDATION TRUST" | OAPs.trusts.costs$Breakdown1Description == "Birmingham And Solihull Mental Health NHS Foundation Trust")
```

### Export

```{r}
write.csv(OAPs.bham.costs, file = "OAPs_bham_costs.csv")
```