---
title: "Out of Area Placements England"
author: "Vanessa Fillis"
date: "08/01/2021"
output: html_document
---

# Out of Area Placements in Mental Health Services in England

## Activate packages

```{r}
library(lubridate)
library(tidyr)
library(dplyr)
```

## Data import

In a previous step I have downloaded data for Out of Area Placements (OAPs) in mental health services between October 2016 and October 2020 (latest available data set) from [NHS Digital](https://digital.nhs.uk/data-and-information/publications/statistical/out-of-area-placements-in-mental-health-services).

```{r}
OAPs <- read.csv("oaps2016to2020.csv", stringsAsFactors = FALSE)
```

## Inspect the data frame

What I can see:

- The column containing the date is not in date format, but as character. I have to add a date column that contains the publication periodm in date format. Because the data frame is so large I will to this *after* I have filtered for the questions I want to look at. 
- The column **value** is in character format as well. I will transform it into numeric values. 

```{r}
head(OAPs)
OAPs$value <- as.numeric(OAPs$value)
head(OAPs)
```

## Filter for inappropriate OAPs in England that are active at the end of the period 

For almost all months, Out of Area Placements is abbreviated. Only in July 2019, the full term has been used. This has to be considered when writing the filter. 

```{r}
OAPs.England.active <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.England.active <- subset(OAPs.England.active, OAPs.England.active$Question == "Inappropriate OAPs active at period end" | OAPs.England.active$Question == "Inappropriate out of area placements active at period end")

OAPs.England.active <- subset(OAPs.England.active, OAPs.England.active$Breakdown1 == "England")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.England.active <- tidyr::separate(data = OAPs.England.active, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.England.active$beginning.of.publication.period <- lubridate::dmy(OAPs.England.active$beginning.of.publication.period)

OAPs.England.active$end.of.publication.period <- lubridate::dmy(OAPs.England.active$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.England.active <- OAPs.England.active[-c(7:11)]
```

### Export

```{r}
write.csv(OAPs.England.active, file = "OAPs_Eng_active.csv")
```

## Filter for total inappropriate OAP days in England for each period

I also want to look at total inappropriate OAP days in England. 

```{r}
OAPs.England.days <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.England.days <- subset(OAPs.England.days, OAPs.England.days$Question == "Total number of inappropriate OAP days over the period" | OAPs.England.days$Question == "Total number of inappropriate out of area placement days over the period")

OAPs.England.days <- subset(OAPs.England.days, OAPs.England.days$Breakdown1 == "England")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.England.days <- tidyr::separate(data = OAPs.England.days, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.England.days$beginning.of.publication.period <- lubridate::dmy(OAPs.England.days$beginning.of.publication.period)

OAPs.England.days$end.of.publication.period <- lubridate::dmy(OAPs.England.days$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.England.days <- OAPs.England.days[-c(7:11)]
```

### Export

```{r}
write.csv(OAPs.England.days, file = "OAPs_Eng_days.csv")
```

## Filter for total inappropriate OAP costs in England for each period

I also want to look at costs caused by inappropriate OAPs each month. 

```{r}
OAPs.England.costs <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.England.costs <- subset(OAPs.England.costs, OAPs.England.costs$Question == "Total Cost For Inappropriate OAPs in Period" | OAPs.England.costs$Question == "Total Cost For Inappropriate out of area placements in Period")

OAPs.England.costs <- subset(OAPs.England.costs, OAPs.England.costs$Breakdown1 == "England")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.England.costs <- tidyr::separate(data = OAPs.England.costs, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.England.costs$beginning.of.publication.period <- lubridate::dmy(OAPs.England.costs$beginning.of.publication.period)

OAPs.England.costs$end.of.publication.period <- lubridate::dmy(OAPs.England.costs$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.England.costs <- OAPs.England.costs[-c(7:11)]
```

### Export

```{r}
write.csv(OAPs.England.costs, file = "OAPs_Eng_costs.csv")
```


## How many receiving providers are private in October 2020?

I want to look at where patients are being sent and if the receiving providers are private or NHS organisations. 

```{r}
OAPs.receivers <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.receivers <- subset(OAPs.receivers, OAPs.receivers$Question == "Inappropriate OAPs active at period end" | OAPs.receivers$Question == "Inappropriate out of area placements active at period end")

OAPs.receivers <- subset(OAPs.receivers, OAPs.receivers$Breakdown1 == "ReceivingProvider")
```

### Convert date column

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.receivers <- tidyr::separate(data = OAPs.receivers, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.receivers$beginning.of.publication.period <- lubridate::dmy(OAPs.receivers$beginning.of.publication.period)

OAPs.receivers$end.of.publication.period <- lubridate::dmy(OAPs.receivers$end.of.publication.period)
```

### Delete empty columns

```{r}
OAPs.receivers <- OAPs.receivers[-c(9:11)]
```

### Filter for October 2020

```{r}
OAPs.receivers.oct20 <- subset(OAPs.receivers, OAPs.receivers$beginning.of.publication.period == "2020-10-01")
```

### Export

```{r}
write.csv(OAPs.receivers.oct20, file = "OAPs_receivers_oct20.csv")
```


